Transfinite Games for Countable Nondeterminism

### 2017-03-08 nondeterminism/nondeterminism.tex
Getting there.
Arrrgh!

\newcommand{\stle}{{\;\le_s\;}}
\newcommand{\exle}{\sqsubseteq}
\newcommand{\exlub}{\bigsqcup}
\newcommand{\dv}{{\text{\lightning}}}
\DeclareMathOperator{\pocl}{pocl}
\newcommand{\opto}{\longrightarrow}

@@

  We have a nondeterministic strategy $?$ for $\bN\cong I\implies\bN$ where the player nondeterministically chooses any number - so $? = P_{\bN}^P$ as sets of positions.  $?$ is compact because all plays in $?$ have length at most $2$.

  When we compose these strategies, the resulting strategy $\comp\sigma?$ contains the infinite chain of positions $qq,qqaq,qqaqaq,qqaqaqaq,\dots$, but does not contain any limiting plays.  So it is not compact.  It is, however, locally compact, since it is the union of compact strategies $\sigma_n$, where $\sigma_n$ has the unique maximal play
  \[
    q\underbrace{qa\dots qa}_{\textrm{$n$ times}}a
    \]
\end{example}

@@

The numerals $\underline{n}$ will each be interpreted as the strategy for $\bN$ that always returns the value $n$.  The nondeterministic oracle $?$ will be interpreted as the strategy $?$ given at the level of trace sets by $?=P_{\bN}^P$.  None of these strategies have any divergent positions.

@@

\begin{proposition}
  Let $A,B,C$ be games, let $\sigma,\sigma'$ be strategies for $A\implies B$ and let $\tau,\tau'$ be strategies for $B\implies C$.  Suppose that $\sigma\stle\sigma'$ and $\tau\stle\tau'$.  Then:
  \[
    \comp\tau\sigma\stle\comp{\tau'}{\sigma'}
    \]
  \begin{proof}
    First suppose that $\tau=\tau'$.

@@

\begin{proposition}
  Let $A,B$ be games and let $\sigma,\tau$ be strategies for $A\implies B$ such that $\sigma\stle\tau$.  Then $\oc\sigma\stle\oc\tau$ as strategies for $\oc A\implies \oc B$.
  \begin{proof}
    If $s\in\oc\sigma$, then $s\vert_{A_n,B_n}\in\sigma$ for all $n$ and therefore $s\vert_{A_n,B_n}\in\sigma'$ for all $n$.  It follows that $\oc\sigma\subset\oc\tau$.

@@

\subsection{Soundness}

@@

\begin{lemma}
  \label{jimsLemma}
  Let $M$ be a reducible program of type $T$.  If $M\opto N$ then $\deno{N}\exle\deno{M}$.
  \begin{proof}
    This is easy to show inductively by analysing the reduction rules.
  \end{proof}
\end{lemma}

\begin{theorem}[Soundness]
  Let $M$ be a program of type $\nat$.  If $M\mustconverge$, then $D_{\deno{M}} = \emptyset$.
  \begin{proof}
    By Lemma \ref{jimsLemma}, 
\end{theorem}

\subsection{Adequacy}

I don't think I'll have a full abstraction result yet, but I could add a section explaining how this model gives us some progress towards full abstraction.

>>>>

\newcommand{\stle}{{\;\le_s\;}}
\newcommand{\exle}{\sqsubseteq}
\newcommand{\exlub}{\bigsqcup}
\newcommand{\dv}{{\text{\lightning}}}
\DeclareMathOperator{\pocl}{pocl}
\newcommand{\opto}{\longrightarrow}
\newcommand{\n}{{\mathtt{n}}}
\newcommand{\IfO}{{\mathtt{If0}\;}}
\newcommand{\suc}{{\mathtt{suc\;}}}
\newcommand{\0}{{\mathtt{0}}}
\newcommand{\plot}{\triangleleft}

@@

  We have a nondeterministic strategy $\wn$ for $\bN\cong I\implies\bN$ where the player nondeterministically chooses any number - so $\wn = P_{\bN}^P$ as sets of positions.  $\wn$ is compact because all plays in $\wn$ have length at most $2$.

  When we compose these strategies, the resulting strategy $\comp\sigma\wn$ contains the infinite chain of positions $qq,qqaq,qqaqaq,qqaqaqaq,\dots$, but does not contain any limiting plays.  So it is not compact.  It is, however, locally compact, since it is the union of compact strategies $\sigma_n$, where $\sigma_n$ has the unique maximal play
  \[
    q\underbrace{qa\dots qa}_{\textrm{$n$ times}}a
    \]
\end{example}

@@

The numeral $\0$ will each be interpreted as the strategy for $\bN$ that always returns the value $0$, and the function $\suc$ will be the strategy for $\bN\iimpl\bN$ that interrogates its argument and then returns a value that is one greater.  The nondeterministic oracle $\wn$ will be interpreted as the strategy $\wn$ given at the level of trace sets by $\wn=P_{\bN}^P$.  None of these strategies have any divergent positions.

@@

\begin{proposition}
  \label{compositionMonotonicSO}
  Let $A,B,C$ be games, let $\sigma,\sigma'$ be strategies for $A\implies B$ and let $\tau,\tau'$ be strategies for $B\implies C$.  Suppose that $\sigma\stle\sigma'$ and $\tau\stle\tau'$.  Then:
  \[
    \comp\tau\sigma\stle\comp{\tau'}{\sigma'}
    \]
  \begin{proof}
    First suppose that $\tau=\tau'$.

@@

\begin{proposition}
  \label{exponentiationMonotonicSO}
  Let $A,B$ be games and let $\sigma,\tau$ be strategies for $A\implies B$ such that $\sigma\stle\tau$.  Then $\oc\sigma\stle\oc\tau$ as strategies for $\oc A\implies \oc B$.
  \begin{proof}
    If $s\in\oc\sigma$, then $s\vert_{A_n,B_n}\in\sigma$ for all $n$ and therefore $s\vert_{A_n,B_n}\in\sigma'$ for all $n$.  It follows that $\oc\sigma\subset\oc\tau$.

@@

Now let $A$ be a game and let $\tau$ be a strategy for $A\iimpl A$.  The set $\Strat(A)$ of strategies for $A$ is directed-complete under the stable ordering $\stle$ and map $\tau_*\from\Strat(A)\to\Strat(A)$ given by $\sigma\mapsto\icomp\tau\sigma$ is monotone with respect to $\stle$, by Propositions \ref{compositionMonotonicSO} and \ref{exponentiationMonotonicSO}.  Therefore, by the Bourbaki-Witt theorem, there is a $\stle$-least fix-point of $\tau_*$, which we shall call $\fix_\tau$.

Let $T$ be a PCF term with denotation $A$ and let $\chi_A$ be the strategy for $((A \iimpl A) \iimpl A) \iimpl ((A \iimpl A) \iimpl A)$ corresponding to the $\lambda$-term $\lambda F.\lambda f. f(F f)$.  We define
\[
  \deno{\Y_T} = \fix_{\chi_A}
  \]

\subsection{Soundness and adequacy}

@@

\begin{lemma}
  \label{jimsLemma}
  Let $M$ be a reducible program of type $T$.  Let $F_M = \{N\suchthat M\opto N\}$.  Then $T_{\deno{M}}\subset\bigcup_{N\in F_M} T_{\deno{N}}$ and $D_{\deno{M}}\subset \bigcup_{N\in F_M} D_{\deno{N}}$.
  \begin{proof}
    Structural induction on $M$.  If $M=\wn$, then $F_M$ is the set of numerals, and we have $T_\wn = \bigcup_{n\in \omega} \{\emptyplay, qn\}$ and $D_\wn=\emptyset$.  Otherwise, write $M = FN$ for some term $L$ of type $S\to T$ and some term $N$ of type $T$.  Write $A=\deno{S}$, $B=\deno{T}$, $\sigma=\deno{N}$ and $\tau=\deno{F}$.  So $\sigma$ is a strategy for $A$ and $\tau$ is a strategy for $A\iimpl B = \oc A \implies B$.  If $FN$ is reducible via one of the rules
    \[
      \begin{tabular}{cc}
        \inferrule{ }{(\lambda x.M)N\opto M[N/x]}
          & \inferrule{ }{\Y M \opto M(\Y M)} \\
        \inferrule{ }{\IfO \0 \opto \lambda x.\lambda y.x}
          & \inferrule{ }{\IfO (\suc\n) \opto \lambda x.\lambda y.y(\n)}
      \end{tabular}
      \]
    then $FN$ is not reducible via any other rule; moreover, in all of these cases, the terms on either side of the $\opto$ arrow have the same denotations (by structural induction in the case of the $\beta$-reduction).  Now suppose that $FN$ is reducible by one of the rules
    \[
      \begin{tabular}{cc}
        \inferrule{M\opto M'}{\suc M\opto \suc M'}
          & \inferrule{M\opto M'}{\IfO M \opto \IfO M'}
      \end{tabular}
      \]
    Let $s\in T_{\comp\tau\sigma}$, and write $s=\s\vert_B$, where $\s\in\sigma\|\tau$.  By structural induction, there is some term $N'$, having denotation $\sigma'$, such that $N\to N'$ and $\s\vert_A\in T_{\sigma'}$.  Then we have $s=\s\vert_B\in T_{\comp\tau{\sigma'}}$.

    Now suppose that $d\in D_{\comp\tau\sigma}$.  Then there is some $\d\in \oc\sigma\dv\tau$ such that $\d\vert_B\prefix d$.  Since the denotations for $\IfO$ and $\suc$ do not contain any divergences, we must have $\d\in T_\sigma$ and $\d\vert_{\oc A}\in D_{\oc \sigma}$.  Moreover, the denotations of $\IfO$ and $\suc$ only interrogate their argument once.  So $\d\vert_{A_0}\in D_\sigma$ and $\d\vert_{A_n}=\emptyplay$ for $n\ne 0$.  By structural induction there is some $N'$, having denotation $\sigma'$, such that $N\opto N'$ and $\d\vert_{A_0}\in D_{\sigma'}$, and therefore we have $\d\vert_A\in D_{\oc\sigma'}$.  It follows that $\d\in D_{\icomp{\tau}{\sigma'}}$.

    Lastly, suppose that $F N$ is reducible by the rule
    \[
      \begin{tabular}{c}
        \inferrule{M\opto M'}{M N\opto M' N}
      \end{tabular}
      \]
    Let $s\in T_{\comp\tau\sigma}$ and write $s=\s\vert_B$, where $\s\in\sigma\|\tau$.  In particular, $\s\in T_\tau$.  By structural induction, there is some term $F'$, having denotation $\tau'$, such that $F\to F'$ and $\s\in T_{\tau'}$.  Then $s=\s\vert_B\in T_{\comp{\tau'}{\sigma}}$.

    Now suppose that $d\in D_{\comp\tau\sigma}$.  Then there is some $\d\in\oc\sigma\dv\tau$ such that $\d\vert_B\prefix d$.  Suppose first that $\d\vert_A\in D_{\oc\sigma}$ and that $\d\in T_\tau$.  By structural induction, there is some $F'$, having denotation $\tau'$, such that $F\to F'$ and $\d\in T_{\tau'}$.  Then $\d\in\oc\sigma\dv\tau'$, so $\d\vert_B\in D_{\comp{\tau'}\sigma}$.

    Suppose instead that $\d\vert_A\in T_{\oc\sigma}$ and that $\d\in D_\tau$.  By structural induction, there is some $F'$, having denotation $\tau'$, such that $F\to F'$ and $\d\in D_{\tau'}$.  Then $\d\in\oc\sigma\dv\tau'$, so $\d\vert_B\in D_{\tau'}$.  
  \end{proof}
\end{lemma}

\begin{theorem}[Consistency]
  Let $M$ be a program of type $\nat$.  If $M\mustconverge$, then $D_{\deno{M}} = \emptyset$.
  \begin{proof}
    Suppose $d\in D_{\deno M}$.  The only irreducible programs of type $\nat$ are the numerals, whose denotations do not contain any divergences, so $M$ must be reducible.  Therefore, by Lemma \ref{jimsLemma}, there exists some $M_1$ such that $M\to M_1$ and $d\in D_{\deno {M_1}}$.  So $M_1$ is reducible and we may apply Lemma \ref{jimsLemma} again to get a term $M_2$ such that $M_1\to M_2$ and $d\in D_{\deno {M_2}}$.  We keep on applying Lemma \ref{jimsLemma} and build up a sequence $M\to M_1\to M_2\to M_3\to\cdots$.  Therefore, $M\not\mustconverge$.
  \end{proof}
\end{theorem}

To prove computational adequacy, we define Plotkin-style approximation relations (\cite{PlotkinApRels}, cited in \cite{LairdOrdinalGames}).

\begin{definition}
  For each PCF type $T$, we define a relation $\plot_T$ between strategies for $\deno{T}$ and closed terms of type $T$:

  \begin{itemize}
    \item $\sigma \plot_\nat M$ if $M\not\mustconverge$ implies that $q\in D_\sigma$ and $qn\in T_\sigma$ implies that $M\opto^* \n$.
    \item $\tau \plot_{S\to T} P$ if $\sigma\plot_S N$ implies that $\comp\tau\sigma \plot_T P N$
  \end{itemize}
\end{definition}

\begin{lemma}
  \begin{enumerate}[i)]
    \item If $M$ is a term of type $\nat$ such that $M\opto N$ for some unique term $N$, and if $\sigma\plot_\nat N$, then $\sigma\plot_\nat M$
    \item If $M$ is a term of type $T$ and $\sigma,\tau$ are strategies for $\deno{M}$ such that $\sigma\stle\tau$, and if $\tau\plot_TM$, then $\sigma\plot_TM$.
    \item If $M$ is a term of type $T$ and $F$ is a stably directed set of strategies for $\deno{M}$ such that $\sigma\plot_TM$ for all $\sigma\in F$, then $\exlub F\plot_TM$.  
  \end{enumerate}
  \begin{proof}
    \begin{enumerate}[(i):]
      \item Suppose $M\opto N$ for some unique term $N$ and suppose that $\sigma\plot_\nat N$.  Then, if $M\not\mustconverge$, we must have $N\not\mustconverge$.  Therefore, $q\in D_\sigma$.  Suppose $qn\in T_\sigma$.  Then $N\opto^*\n$, so $N\opto^*\n$.
      \item Structural induction on $T$.  let $M$ be a term of type $\nat$.  Suppose $\sigma,\tau$ are strategies for $\bN$, that $\sigma\stle\tau$ and that $\tau\plot_T M$.  If $M\not\mustconverge$, then $q\in D_\tau$, so $q\in D_\sigma$.  If $qn\in T_\sigma$, then $qn\in T_\tau$, so $M\opto^*\n$.

        Now let $P$ be a term of type $S\to T$.  Write $A=\deno{S}$ and $B=\deno{T}$.  Suppose that $\tau,\tau'$ are strategies for $A\iimpl B$ such that $\tau\stle\tau'$ and such that $\tau'\plot_{S\to T}P$.  Suppose that $\sigma\plot_SN$ for some term $N$ of type $S$.  Then $\icomp{\tau'}{\sigma}\plot_TPN$.  Since $\tau\stle\tau'$, $\icomp{\tau}{\sigma}\stle\icomp{\tau'}{\sigma}$, by Propositions \ref{compositionMonotonicSO} and \ref{exponentiationMonotonicSO}.  Therefore, by structural induction, we have $\icomp{\tau}{\sigma}\plot_TPN$.  Since $\sigma,N$ were abitrary, it follows that $\tau\plot_{S\to T}P$.
      \item Structural induction on $T$.  Let $M$ be a term of type $\nat$ and let $F$ be a stably directed set of strategies for $\bN$ such that $\sigma\plot_TM$ for all $\sigma\in F$.  Suppose $M\not\mustconverge$.  Then $q\in D_\sigma$ for all $\sigma\in F$, so $q\in D_{\exlub F}$.  Now suppose that $qn\in T_{\exlub F}$.  Then $qn\in T_\sigma$ for some $\sigma\in F$, and so $M\opto^*\n$.

        Now let $P$ be a term of type $S\to T$.  Write $A=\deno{S}$ and $B=\deno{T}$.  Suppose that $F$ is a stably directed set of strategies for $A\iimpl B$ such that $\tau\plot_{S\to T}P$ for all $\tau\in F$.  Let $N$ be a term of type $S$ and let $\sigma$ be a strategy for $A$ such that $\sigma\plot_SN$.  Then $\icomp\tau\sigma\plot_TPN$ for all $\tau\in F$.  By structural induction, $\exlub_{\tau\in F}(\icomp\tau\sigma)\plot_TPN$.  Now 
        \qedhere
    \end{enumerate}
  \end{proof}
\end{lemma}
